---
title: "Exercise Sheet 7"
author: "Maiwen Caudron-Herger ; Carl Herrmann"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

We will work with a dataset of tumor expression data of patients with acute myeloid leukemia (AML) or acute lymphocytic leukemia (ALL):

```{r, echo = FALSE}
all.aml = read.delim('https://www.dropbox.com/scl/fi/q4rbpl3un6oqbbwoxe1qw/all.aml.cleaned.csv?rlkey=igwbna22d6hpdmw3os79zoy7z&dl=1',header=TRUE)
all.aml.anno = readRDS(url("https://www.dropbox.com/scl/fi/0lqjh3jcskki2bb8hxwj6/all.aml.anno.rds?rlkey=cijsx39kr9gx9bjl6wzpmkgnz&dl=1"))
#
# we convert all.aml into a data matrix rather than a data.frame
all.aml = data.matrix(all.aml)
i.all = which(all.aml.anno$ALL.AML=='AML') ## we extract the row numbers of the ALL patients
```

# 0. Introduction and Objectives

In the previous Exercise Sheet we have looked at **hypothesis testing**, learned how the process works and then learned about the t-test function.

In this Exercise Sheet you will learn more about hypothesis testing, specifically about how to deal with **non-parametric tests** and **proportion testing**.

You should be familiar with both these problems from the lecture.

# 1. Non-parametric test

What if the data is not normally distributed? In that case, we are no supposed to use the t-test for testing differences between mean values! Let us us see an example.

At the beginning, we checked if the distribution of the data in the samples corresponds to a normal distribution.

Check again the distribution of the expression values for the gene FOSB

```{r}
expression = all.aml['FOSB',]
hist(expression,breaks = 30);qqnorm(expression);qqline(expression)
```

This looks everything but normal! In that case, we cannot apply the t.test, but need to apply a **non-parametric test** called the *Wilcoxon test* (check the lecture notes!). This test is performed not on the *values* (like the t-test) but on the *ranks* of these values (remember the difference between the Pearson's and the Spearman's correlations!)

```{r}
exp.all = all.aml['FOSB',i.all]
exp.aml = all.aml['FOSB',-i.all]
wilcox.test(exp.all,exp.aml)
```

Compare the obtained p-value with the p-value obtained if we would have used the t-test:

```{r}
exp.all = all.aml['FOSB',i.all]
exp.aml = all.aml['FOSB',-i.all]
t.test(exp.all,exp.aml)
```

The p-values are very different!! So is the difference of expression between ALL and AML patients for this gene significant or not taking $\alpha=0.05$?

Here, we **cannot** trust the t-test due to the non-normality of the data! Hence, the correct p-value is the one from the Wilcoxon test.

> Check another random gene!


# 2. Proportion tests

The t-test and wilcoxon tests are **tests of the mean**; we are comparing the means of two samples and looking for significant differences.

But there are other hypothesis that one might want to test, related to the relationship between two categorical variables:

-   is the proportion of men **significantly** higher in the patients from Louisa compared to the ones from Buckingham?
-   is the propotion of smothers in under 18 in Germany **significantly** higher than in other european countries?

The proportion test (Fisher Extact Test or chi2 test) are used to investigate the relationship between 2 categorical variables, starting from a **contingency table**. We will use a dataset with clinical informations about breast cancer patients.

```{r}
dat.brca = read.delim(url('https://www.dropbox.com/scl/fi/gy2jnj7o7p1398azkw6b1/gbsg_ba_ca.dat?rlkey=a0j003a1ehtt68usuyd32ta56&dl=1'), stringsAsFactors = FALSE)
```

Check which variables in this dataset are categorical/ordinal/numerical. We can now check if there is a significant relationship between some variables. For example, we can verify if the choice of treatment with tamoxifen (variable `hormon`) is related to the pre-/post-monopausal status (variable `meno`)

First, we can build the contingency table table for these 2 variables:

```{r}
## build contingency table
table(dat.brca$meno,dat.brca$hormon)
```

------------------------------------------------------------------------

We can compute the **odds-ratio (OR)** for these two variables:

```{r}
CT = table(dat.brca$meno,dat.brca$hormon)

OR = (CT[1,1]/CT[1,2])/(CT[2,1]/CT[2,2])
OR
```

> How would the odds-ratio look like if you would transpose the matrix?


Now we can run the one-sided Fisher Exact Test (FET). The H0/H1 hypothesis are:

-   H0: the odds-ratio is not significantly larger than one
-   H1: the odds-ratio is significantly larger than one

```{r}
## build contingency table
tab = table(dat.brca$meno,dat.brca$hormon)
#
## run the FET
fisher.test(tab,alternative = 'greater')
```

Check if your computation of the odds-ratio is right! 


> Run the two-sided Fisher test; first, formulate the hypothesis!
> Check the p-values: what do you observe?


We can also use the **chi-square test** to answer the same question. The Chi-square test compares the **observed** number of occurences in the contingency table to the **expected** number of occurences if there was no relationship between the variables.

-   H0: the observed and expected occurences are not significantly different
-   H1: the observed and expected occurences are significantly different

```{r}
chisq.test(tab)
```

> Is this a one-sided or 2-sided test?

Now we want to verify the impact of age on the grade of the tumor. We categorize the patients in under and over 40 year groups, and perform a chi-squared test:

```{r}
## contingency table
tab = table(dat.brca$age>40,dat.brca$grade)
tab
##
chisq.test(tab)
```

We can determine the table of expected counts:

```{r}
tot = apply(tab,2,sum) # this is the total number of occurences in the 3 categories, independently of age
tot

age = apply(tab,1,sum) # this is the total number of persons above/below 40, independently of grade
age
```

```{r}
tot.proportions = tot/sum(tot)

tab.exp = sapply(tot.proportions,function(x) {x*age})
tab.exp
```

> How would you compute the chi-square test statistic using the `tab` and `tab.exp` tables?

# 3. Power of a test

We have seen that $\alpha$ controls the **False-positive rate**. False positive means seing an difference that is not there...

The other type of error are **False-negatives**, i.e. **not seeing** a difference that actually exists!

Let us imagine the following scenario:

You are working in a clinical laboratory, testing patients for a rare genetic condition that affects enzyme levels in blood. The condition can be detected by measuring the concentration of a specific biomarker.

**Healthy individuals** have an average biomarker level of $\mu_H = 50\,\mu g/L$ with a standard deviation of $\sigma = 12\,\mu g/L$.

**Patients with the condition** have elevated levels: $\mu_D = 58\,\mu g/L$ with the same standard deviation $\sigma = 12\,\mu g/L$.

Due to biological variation, you measure each patient's biomarker level $n=5$ times and use the average of these measurements for diagnosis.

```{r}
mu_healthy = 50    # mean biomarker level in healthy individuals
mu_disease = 58    # mean biomarker level in diseased individuals
sigma = 12         # standard deviation (same for both groups)
n = 5              # number of replicate measurements per patient
alpha = 0.05       # significance level
```

The **null hypothesis H0** is: *"The patient is healthy (does not have the condition)"*

Here are the average biomarker measurements from your first 5 patients today:

```{r}
patient_means = c(56.3, 62.1, 51.8, 49.2, 54.7)
```

## What is the expected distribution of mean biomarker levels?

For healthy individuals, when taking $n=5$ measurements and averaging them, the distribution of sample means is:

```{r}
x = seq(35, 65, by = 0.1)
y = dnorm(x, mean = mu_healthy, sd = sigma/sqrt(n))
plot(x, y, type='l', lwd=3, col='blue',
     xlab='Average biomarker level (μg/L)',
     ylab='Density',
     main=paste0('Distribution of mean biomarker levels in healthy individuals (n=',n,' measurements)'))
```

## When would you reject H0 and diagnose the patient as having the condition?

Using a one-sided test with $\alpha=0.05$:

```{r}
rejection_threshold = qnorm(p = alpha, mean = mu_healthy, sd = sigma/sqrt(n), lower.tail = FALSE)
rejection_threshold
```

If a patient's average biomarker level is **above** this value, you would reject H0 and diagnose them with the condition (**positive** diagnosis). Otherwise, you would not reject H0 (**negative** diagnosis).

```{r}
plot(x, y, type='l', lwd=3, col='blue',
     xlab='Average biomarker level (μg/L)',
     ylab='Density',
     main=paste0('Diagnostic decision rule (n=',n,' measurements)'))
abline(v = rejection_threshold, lwd=3, lty=2, col='red')
abline(v = patient_means, lty=3, lwd=2, col='darkgreen')
legend('topright', 
       legend=c('Healthy distribution','Diagnostic threshold','Patient measurements'),
       col=c('blue','red','darkgreen'), lwd=c(3,3,2), lty=c(1,2,3))
text(rejection_threshold, max(y)*0.8, 
     paste0('Threshold = ', round(rejection_threshold,2)), 
     pos=4, col='red')
```

Based on this threshold, patient 2 would be diagnosed with the condition, while the other 4 patients would be classified as healthy.

## The problem: False negatives in diseased patients

**However**, it turns out that all 5 of these patients actually **DO have the condition**! They were referred from a specialist clinic that had already identified them through genetic testing. So H0 is **false** for **all** of these patients.

This means you failed to detect the disease in 4 out of 5 patients: **80% false-negatives**!

This is the **Type II error rate (β)**, also called the **false-negative rate**. The complement **(1-β)** is called the **power** or **sensitivity** of the test.

## Calculating the false-negative rate (β)

Let's simulate what happens when we test 1000 patients who **actually have the condition**:

```{r}
### Parameters
mu_healthy = 50
mu_disease = 58
sigma = 12
n = 5
alpha = 0.05

### Simulate 1000 patients with the disease (n measurements each)
patient_means = sapply(1:1000, function(i) {
  measurements = rnorm(n, mean = mu_disease, sd = sigma)
  mean(measurements)
})

### Diagnostic threshold for H0: "patient is healthy"
rejection_threshold = qnorm(alpha, mean = mu_healthy, sd = sigma/sqrt(n), lower.tail = FALSE)

### For how many patients did we FAIL to detect the disease?
### H0 is rejected (disease detected) if mean is ABOVE threshold
### H0 is NOT rejected (disease missed) if mean is BELOW threshold
beta = sum(patient_means < rejection_threshold) / length(patient_means)

cat(sprintf("False-negative rate (β) = %.1f%%\n", beta*100))
cat(sprintf("Test power/sensitivity (1-β) = %.1f%%\n", (1-beta)*100))
```

This means our diagnostic test only detects the disease in about 43 % of patients who actually have it. This is a **very poor sensitivity** for a diagnostic test!

## Visualizing Type II errors

```{r}
# Plot both distributions
x = seq(35, 75, by = 0.1)
y_healthy = dnorm(x, mean = mu_healthy, sd = sigma/sqrt(n))
y_disease = dnorm(x, mean = mu_disease, sd = sigma/sqrt(n))

plot(x, y_healthy, type='l', lwd=3, col='blue',
     xlab='Average biomarker level (μg/L)',
     ylab='Density',
     main='Understanding false-negatives in diagnostic testing',
     ylim=c(0, max(c(y_healthy, y_disease))))
lines(x, y_disease, lwd=3, col='purple')
abline(v = rejection_threshold, lwd=3, lty=2, col='red')

# Shade the false-negative region
x_fn = x[x < rejection_threshold]
y_fn = dnorm(x_fn, mean = mu_disease, sd = sigma/sqrt(n))
polygon(c(x_fn, rev(x_fn)), c(y_fn, rep(0, length(y_fn))), 
        col=rgb(0.5,0,0.5,0.3), border=NA)

legend('topright', 
       legend=c('Healthy patients','Diseased patients','Diagnostic threshold','False-negative region'),
       col=c('blue','purple','red',rgb(0.5,0,0.5,0.3)), 
       lwd=c(3,3,3,10), lty=c(1,1,2,1))
```

The **purple shaded area** represents diseased patients whose biomarker levels fall below the threshold—these are the **false-negatives** that we fail to diagnose.


> How does the false-negative rate β change with increasing number of measurements per patient (i.e., increasing n)? Try n = 5, 10, 20, 50.

> Try to find a combination of:
> - Effect size (difference between μ_healthy and μ_disease)
> - Significance level (α)
> - Number of measurements (n)
> 
> ...so that the test has at least 80% power (β < 20%) while maintaining α = 0.05.


------------------------------------------------------------------------

# Exercises

#### Exercise 1

1.  Check the expression of a random gene in the ALL/AML dataset. Are the values normally distributed?
2.  Run both tests on this gene. Is there a difference between p-values? What did you expect?
3.  Write a code that runs both tests on every gene and registers each p-value in a vector.
4.  Do a scatter plot of -log10(Pvalue of t-test) vs. -log10(Pvalue of wilcoxon test); are there strong deviations?

#### Exercise 2

What test would you use for the following questions?

-   A lotion company has to figure out whether their last product is more likely to give men acne rather than females
-   The department of education wants to find out whether social science students have higher grades than science students
-   A biologist needs to find out whether a specific gene is more likely to be silenced in lactose intolerant people.

#### Exercise 3

A pharmaceutical company has a new corona vaccine that works on 178 out of 200 patient. 75% of the batch of patients had been previously vaccinated against corona with another previous vaccine, and on 143 of these did the vaccine work. Is there a significant disproportion to say that the vaccine has a higher chance of working on people who are already vaccinated against corona if our p-value threshold is 0.05?

------------------------------------------------------------------------

# Going further

What does it mean, when we say that we cannot trust the t-test when the data is not normaly distributed? Remember that the significance value $\alpha$ represents the **false-positive rate (FPR)**. Hence, with an $\alpha$ of 5%, if $H_0$ is true, we have a 5% risk of rejecting $H_0$ (False-positive).

Let us suppose that the $H_0$ Hypothesis: "The expectation values of two distributions are equal" holds. For example, we can generate 2 random samples, drawn from the same distribution, and perform a t-test. Here, $H_0$ would hold! What would the p-value be? Well, if we perform this 10000 times, we would obtain a uniform distribution:

```{r}
set.seed(123)
alpha = 0.05
p = sapply(1:10000,function(i) {
  x = rnorm(10);y = rnorm(10)
  t.test(x,y)$p.value
})
hist(p,breaks=20);abline(v=alpha,col='red',lwd=3)
```

Approximately 5% of these p-values are smaller than $\alpha$ (see red line):

```{r}
sum(p<alpha)/length(p)
```

These would be **False positives**. So here, we indeed have $\alpha$ = False-positive rate!

This also holds true for different values of $\alpha$:

```{r}
alphas = c(0.01,0.02,0.03,0.04,0.05,0.07,0.1,0.2)
fpr = sapply(alphas,function(alpha) {
  sum(p<alpha)/length(p)
})
plot(alphas,fpr,xlab='alpha',pch=19,col='red');abline(0,1,lwd=2,lty=2,col='lightgrey')
```

Now what if the distributions of values had same expectation values, but were not normaly distributed? Let's sample x and y from a t-Distribution, instead of a normal distribution:

```{r}
p = sapply(1:10000,function(i) {
  x = rt(10,df=1);y = rt(10,df=1)
  t.test(x,y)$p.value
})
hist(p,breaks=20);abline(v=alpha,col='red',lwd=3)
```

No longer a uniform distribution!! How often would we reject $H_0$ here?

```{r}
sum(p<alpha)/length(p)
```

Not exactly $\alpha$ anymore... What if we user different values of $\alpha$?

```{r}
alphas = c(0.01,0.02,0.03,0.04,0.05,0.07,0.1,0.2)
fpr = sapply(alphas,function(alpha) {
  sum(p<alpha)/length(p)
})
plot(alphas,fpr,xlab='alpha',pch=19,col='red');abline(0,1,lwd=2,lty=2,col='lightgrey')
```

So in the case of a non-normal distribution of the data, the false-positive rate (FPR) is no longer equal to the significance level $\alpha$... **Hence, we can no longer control the FPR with** $\alpha$!