---
title: "T-tests"
author: "Carl Herrmann, Maiwen Caudron-Herger"
date: "01/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to the concept of t-test

We simulate two datasets:

```{r}
set.seed(323)
ma = 4
mb = 5
s = 3
d = runif(10,0,s)
a = c(ma+d,ma-d)
b = c(mb+d,mb-d)
```

```{r}
a
b
```


Computing the mean of both datasets:

```{r}
mean(a);mean(b)
```

```{r}
plot(c(1,2),c(mean(a),mean(b)),pch=19,col='red',xlim=c(0,3),ylim=c(0,10),
     xaxt='n',xlab='',ylab='')
```



```{r}
plot(c(1,2),c(mean(a),mean(b)),pch=19,col='red',xlim=c(0,3),ylim=c(0,10),
     xaxt='n',xlab='',ylab='')
points(c(rep(1,length(a)),rep(2,length(b))),c(a,b),pch=20)
```

```{r}
tt = t.test(a,b,var.equal = TRUE)
tt
```

What is the H0 distribution? t-distribution with 20+20-2 = 38 degrees of freedom


```{r}
x = seq(-5,5,by=0.1)
y = dt(x,df=tt$parameter)
plot(x,y,type='l',lwd=3,col='red');abline(v=tt$statistic,lwd=2,lty=3)
```



## Comparing cholesterol

```{r}
dat = read.delim('https://tinyurl.com/y4fark9g', stringsAsFactors = FALSE)
```

```{r}
chol.m = dat$chol[dat$gender=='male' & !is.na(dat$chol)]
chol.f = dat$chol[dat$gender=='female' & !is.na(dat$chol)]
```

```{r}
boxplot(chol.m,chol.f)
```

```{r}
mean(chol.m)
mean(chol.f)
```

### Question: Do men have a higher cholesterol level compared to women?

-   H0: men have a cholesterol level **lower or equal** than women
-   H1: men have a cholesterol level **higher** than women

```{r}
tt = t.test(chol.m,chol.f,alternative = 'greater',var.equal = TRUE)
tt
```

Let us look at the null distribution

```{r}
df = 400
x = seq(-5,5,by=0.01)
y = dt(x,df=df)
plot(x,y,type='l',lwd=3);abline(v=tt$statistic,lwd=2,col='blue',lty=3)
```

We can obtain the p-value using the `pt` function:

```{r}
pt(tt$statistic,df = length(chol.m)+length(chol.f)-2,lower.tail = FALSE)
```

Would it make a difference if we had used the SNV as a null distribution instead?

```{r}
pnorm(tt$statistic,lower.tail = FALSE)
```

Not really, as the number of data points is high enough...

### Question: Do men have a lower cholesterol level compared to women?

-   H0: men have a cholesterol level **higher or equal** than women
-   H1: men have a cholesterol level **lower** than women

```{r}
tt = t.test(chol.m,chol.f,alternative = 'less',var.equal = TRUE)
tt
```

Let us look at the null distribution

```{r}
df = 400
x = seq(-5,5,by=0.01)
y = dt(x,df=df)
plot(x,y,type='l',lwd=3);abline(v=tt$statistic,lwd=2,col='blue',lty=3)
```

### Question: Do men have a different cholesterol level compared to women?

-   H0: men have a cholesterol level **equal** to that of women
-   H1: men have a cholesterol level **different** than that women

```{r}
tt = t.test(chol.m,chol.f,alternative = 'two.sided',var.equal = TRUE)
tt
```

Let us look at the null distribution

```{r}
df = 400
x = seq(-5,5,by=0.01)
y = dt(x,df=df)
plot(x,y,type='l',lwd=3);abline(v=c(-abs(tt$statistic),abs(tt$statistic)),lwd=2,col='blue',lty=3)
```

## Comparing height

```{r}
height.m = dat$height[dat$gender=='male']
height.f = dat$height[dat$gender=='female']
```

### Question: Do men have a higher height compared to women?

-   H0: men have a height **lower or equal** than women
-   H1: men have a height **higher** than women

```{r}
tt = t.test(height.m,height.f,alternative = 'greater',var.equal = TRUE)
tt
```

Let us look at the null distribution

```{r}
df = 396
x = seq(-20,20,by=0.01)
y = dt(x,df=df)
plot(x,y,type='l',lwd=3);abline(v=tt$statistic,lwd=2,col='blue',lty=3)
```

### Question: Do men have a lower height compared to women?

-   H0: men have a height **higher or equal** than women
-   H1: men have a height **lower** than women

```{r}
tt = t.test(height.m,height.f,alternative = 'less',var.equal = TRUE)
tt
```

Let us look at the null distribution

```{r}
df = 396
x = seq(-20,20,by=0.01)
y = dt(x,df=df)
plot(x,y,type='l',lwd=3);abline(v=tt$statistic,lwd=2,col='blue',lty=3)
```

### Question: Do men have a different height compared to women?

-   H0: men have a height **equal** to that of women
-   H1: men have a height **different** than that women

```{r}
tt = t.test(height.m,height.f,alternative = 'two.sided',var.equal = TRUE)
tt
```

Let us look at the null distribution

```{r}
df = 400
x = seq(-20,20,by=0.01)
y = dt(x,df=df)
plot(x,y,type='l',lwd=3);abline(v=c(-abs(tt$statistic),abs(tt$statistic)),lwd=2,col='blue',lty=3)
```

## Distribution of p-values under H0

We sample 2 datasets $x_i$ and $y_i$ **FROM THE SAME** normal distribution; when performing a t.test to compare the means of these 2 datasets, we know that the H0 hypothesis "The expectation values of the underlying random variables X and Y are equal". What kind of p-value would the test give us??

```{r}
p = sapply(1:10000,function(i) {
  n1=10;n2=15
  x1 = rnorm(n1)
  x2 = rnorm(n2)
  #x2 = rnorm(n2,mean=2)
  
  
##
  t.test(x1,x2)$p.value
})
head(p)
```

```{r}
alpha = 0.05
hist(p,breaks=15);abline(v=alpha,lwd=4,col='red')
sum(p > alpha)/length(p)
```