---
title: "Plenum 05.12.2024"
author: "Carl Herrmann, Maïwen Caudron-Herger"
output: html_document
---

# 1. Non-parametric tests

We consider 2 samples with a **non-normal** distribution; for example, we can sample $x1$ and $x2$ from a Gamma-distribution:

```{r}
set.seed(122)
n1 = 8
n2 = 12
x1 = rgamma(n1,shape=2,rate=0.5)
x2 = rgamma(n2,shape=2,rate=0.21)

##

x = seq(0,20,by=0.05)
y1 = dgamma(x,shape=2,rate=0.5)
y2 = dgamma(x,shape=2,rate=0.21)

plot(x,y1,lwd=3,col='red',type='l');lines(x,y2,lwd=3,col='blue',type='l');points(x1,rep(0.05,n1),col='red',pch=19);points(x2,rep(0.1,n2),col='blue',pch=19)

```

The data is not normally distributed, so **we cannot trust the p-value** which is given by the t-test!

Wilcoxon-test

```{r}
wilcox.test(x1,x2)
```

Let's compute the Wilcoxon test by hand

We have 2 samples, and want to test the difference of the mean using a **two-tailed test**:

```{r}
x = c(x1,x2)
probe = c(rep(1,n1),rep(2,n2))
x;probe
```

```{r}
R1 = sum(rank(x)[probe == 1])
R2 = sum(rank(x)[probe == 2])
R1;R2
```

Now let's compute the U value from the Wilcoxon test

```{r}
U1 = R1-n1*(n1+1)/2
U2 = R2-n2*(n2+1)/2
U1;U2
```

```{r}
U = U1
U
```

We need to look into a significance table for the corresponding critical values, for example [here](https://www.real-statistics.com/statistics-tables/mann-whitney-table/)

We have n1=8 and n2=12 and U=13; looking into the table, the critical values are the following:

-   $\alpha$ = 0.05 : 22
-   $\alpha$ = 0.01 : 15
-   $\alpha=0.005$ : 12
-   $\alpha$ = 0.001 : 7

The observed U value (=`U`) is smaller that the critical value for \$\alpha=\$0.01, but larger than the critical value for the \$\alpha=\$0.005 Hence, we can say that for $\alpha=0.01$, the two-tailed test is significant, but not for $\alpha=0.005$!

We can check this using the `wilcox.test` function:

```{r}
wilcox.test(x1,x2)
```

### Difference wilcox/t.test

```{r}
x = c(1,1.2,1.3,1.4)
y = c(1.5,1.45,1.6,1.7)
plot(rep(1,length(x)),x,xlim=c(0,3),ylim=c(0,2),pch=19,col='red')
points(rep(2,length(y)),y,pch=19,col='blue')

t.test(x,y);wilcox.test(x,y)
```

```{r}
x = c(1,1.2,1.3,1.4)
y = c(4.5,4.45,4.6,4.7)
plot(rep(1,length(x)),x,xlim=c(0,3),ylim=c(0,5),pch=19,col='red')
points(rep(2,length(y)),y,pch=19,col='blue')
t.test(x,y);wilcox.test(x,y)
```

Wilcoxon is based on **ranks**.

## Gender Bias in incomes

```{r}
income = c(0 ,	400 ,	500 ,	550 ,	600 	,650 ,	750 ,	800 ,	900 ,	950 ,	1000 	,1100 ,	1200 ,	1500 ,	1600 ,	1800, 	1900 ,	2000, 	2200, 	3500 )
gender = c('M', 	'W' ,	'M' ,	'W' ,	'M', 	'W', 	'M', 	'M', 	'W', 	'W', 	'M', 	'M', 	'W', 	'M', 	'W', 	'M', 	'M', 	'M', 	'M', 	'M' )
```

```{r}
income_w = income[gender=='W']
income_m = income[gender=='M']
length(income_m);length(income_w)
```

```{r}
wilcox.test(income_w,income_m)
```

```{r}
n=grep('W',gender)
U = sum(n)-length(n)*(length(n)+1)/2
U
```

# 2. Proportion tests

```{r}
dat = read.delim('https://tinyurl.com/y4fark9g', stringsAsFactors = FALSE)
```

What are the categorical variables?

```{r}
View(dat)
```

location / gender / frame

For example, we could test if there is a relationship between the gender of the patients and where they come from:

```{r}
T = table(dat$location,dat$gender)
T
```

This is a **contingency table**, from which we can compute the **odds-ratio**:

$$
\frac{(M_L/F_L)}{(M_B/F_B)}
$$ where $M/L$ represents the number of male/female patients.

```{r}
OR = (T[2,2]/T[2,1])/(T[1,2]/T[1,1])
OR
```

The Null hypothesis, in the case of a two-sided test, is:

H0: the OR is equal to 1 H1: the OR is NOT equal to 1

Let's test this!

```{r}
fisher.test(T)
```

With a p-value of 0.6872, we CANNOT reject the H0 hypothesis, hence we would conclude that there is no gender bias here.

Is there a bias regarding the size of the patient coming from Louisa or Birmingham? Size is not a categorical variable, but we can define a new variable oversized:

```{r}
oversized = dat$height > 69
head(oversized)
```

and compute the contingency table

```{r}
T1 = table(dat$location,oversized)

T1 = T1[c(2,1),]
T1
```

Let's compute again the odds ratio

```{r}
OR = (T1[2,2]/T1[2,1])/(T1[1,2]/T1[1,1])  # (30/171)/ (49/148)
OR
```

We can now try a one-sided test, which would be

-   H0: the ratio of oversized patients from Buckingham is **smaller or eqal** that this ratio from Louisa

-   H1: the ratio of oversized patients from Buckingham is **significantly larger** that this ratio from Louisa

```{r}
fisher.test(T1,alternative = 'greater')
```

We can also perform a chi-squared test here:

```{r}
chisq.test(T1)
```

# 3. Test errors (false-positives, false negatives,...)

## Case 1: H0 is valid

Important validation for a statistical test: The distribution of p-values when H0 is valid should be a uniform distribution. If not, there is something wrong with your test:

We have said that the t-test cannot be applied, if the data is NOT normally distributed...

What does it mean, when we say that we cannot trust the t-test? Remember that the significance value $\alpha$ represents the **false-positive rate (FPR)**. Hence, with an $\alpha$ of 5%, if $H_0$ is true, we have a 5% risk of rejecting $H_0$ (False-positive).

Let us suppose that the $H_0$ Hypothesis: "The expectation values of two distributions are equal" holds. For example, we can generate 2 random samples, drawn from the same distribution, and perform a t-test. Here, $H_0$ would hold! What would the p-value be? Well, if we perform this 10000 times, we would obtain a uniform distribution:

```{r}
set.seed(123)
alpha = 0.05
p = sapply(1:10000,function(i) {
  x = rnorm(100);y = rnorm(100)
  t.test(x,y)$p.value
})
hist(p,breaks=20);abline(v=alpha,col='red',lwd=3)
```

Approximately 5% of these p-values are smaller than $\alpha$ (see red line):

```{r}
sum(p<alpha)/length(p)
```

These would be **False positives**. So here, we indeed have $\alpha$ = **False-positive rate!**

This also holds true for different values of $\alpha$:

```{r}
alphas = c(0.01,0.02,0.03,0.04,0.05,0.07,0.1,0.2)
fpr = sapply(alphas,function(alpha) {
  sum(p<alpha)/length(p)
})
plot(alphas,fpr,xlab='alpha',pch=19,col='red');abline(0,1,lwd=2,lty=2,col='lightgrey')
```

Now what if the distributions of values had same expectation values, but were not normal?? Let's sample x and y from a t-Distribution, instead of a normal distribution:

```{r}
p = sapply(1:10000,function(i) {
  x = rt(10,df=2);y = rt(10,df=2)
  t.test(x,y)$p.value
})
hist(p,breaks=20);abline(v=alpha,col='red',lwd=3)
```

No longer a uniform distribution!! How often would we reject $H_0$ here?

```{r}
sum(p<alpha)/length(p)
```

Not exactly $\alpha$ anymore... What if we user different values of $\alpha$?

```{r}
alphas = c(0.01,0.02,0.03,0.04,0.05,0.07,0.1,0.2)
fpr = sapply(alphas,function(alpha) {
  sum(p<alpha)/length(p)
})
plot(alphas,fpr,xlab='alpha',pch=19,col='red');abline(0,1,lwd=2,lty=2,col='lightgrey')
```

So in the case of a non-normal distribution of the data, the false-positive rate (FPR) is no longer equal to the significance level $\alpha$... **Hence, we can no longer control the FPR with** $\alpha$!

What about Wilcoxon-test?

```{r}
p = sapply(1:50000,function(i) {
  x = rt(10,df=1);y = rt(10,df=1)
  wilcox.test(x,y)$p.value
})
hist(p,breaks=20);abline(v=alpha,col='red',lwd=3)
```

```{r}
sum(p<alpha)/length(p)
```

```{r}
alphas = c(0.01,0.02,0.03,0.04,0.05,0.07,0.1,0.2)
fpr = sapply(alphas,function(alpha) {
  sum(p<alpha)/length(p)
})
plot(alphas,fpr,xlab='alpha',pch=19,col='red',main='Wilcoxon');abline(0,1,lwd=2,lty=2,col='lightgrey')
```

## Case 2: H0 is NOT valid

We run 5000 tests over samples coming from 2 normal distributions with difference in expectation value getting larger:

```{r}
d = c(0.01,0.02,0.05,0.1,0.2,0.5,0.75,1,2)
n = 50
P = lapply(d,function(dd) {
  sapply(1:5000,function(i) {
    x = rnorm(n)
    y = rnorm(n,mean=dd)
    p = t.test(x,y)$p.value
    #p = wilcox.test(x,y)$p.value
    return(p)
  })
})
```

```{r}
lapply(1:length(d),function(i) {hist(P[[i]],main=paste0("Diff in mean: ",d[i]),breaks=20,xlab='P values');abline(v=0.05,col='red',lwd=3)})
```

The False-negatives are the cases to the **right** of the red line!

We can compute the false-negative rate (FNR):

```{r}
sapply(1:length(d),function(i) {sum(P[[i]]>0.05)/length(P[[i]])})
```

## Case 3: mixture of H0/H1

We simulate a gene expression matrix between 10 normal samples and 10 disease samples, for 1000 genes; the H0 hypothesis is: "the expression of the gene has same expectation value in the 2 groups"

-   900 genes show no difference between the 2 groups (H0 is valid)
-   100 genes show a difference (H0 is NOT valid)

```{r}
## 900 genes with no differential expression
set.seed(123)
H0 <- t(sapply(1:900,function(i) {
  x <- rnorm(10)
  y <- rnorm(10)
  c(x,y)
}))

## 100 genes which show differential expression

H1 <- t(sapply(1:100,function(i) {
  x <- rnorm(10)
  y <- rnorm(10,mean=1)
  c(x,y)
}))

## global expression matrix
X <- rbind(H0,H1)
```

Let's display this matrix:

```{r}
library(RColorBrewer)
library(gplots)
breaks <- seq(-3,3,length.out=100)
col <- colorRampPalette(brewer.pal(10,'Spectral'))(length(breaks)-1)
label <- c(rep('red',900),rep('green',100))

heatmap.2(X,col=col,density="none",trace="none",
          breaks=breaks,dendrogram='none',
          labRow ="",Rowv = FALSE,Colv = FALSE,RowSideColors = label)

```

Now, we will compute the t-test for each gene, comparing the 2 group:

```{r}
## compute p-values
p <- apply(X,1,function(x) {
  t.test(x[1:10],x[11:20])$p.value
})
```

Let's display the distribution of the p-values:

```{r}
par(mfrow=c(3,1),mar=c(2,2,2,2))
hist(p[1:900],breaks=30);hist(p[901:1000],breaks=30);hist(p,breaks=30)
```

### Computing false-positives / false-negatives, etc...

We set alpha = 0.05, and determine in which case we reject H0 (red), and in which case we do not (yellow):

```{r}
alpha <- 0.05

label <- rep('yellow',1000)
label[p<=alpha] <- 'red'

heatmap.2(X,col=col,density="none",trace="none",
          breaks=breaks,dendrogram='none',
          labRow ="",Rowv = FALSE,Colv = FALSE,RowSideColors = label)

```

```{r}
## false positives
fp <- sum(p[1:900] <= alpha)
fp
```

```{r}
## true negatives
tn <- sum(p[1:900] > alpha)
tn
```

```{r}
## true positives
tp <- sum(p[901:1000] <= alpha)
tp
```

```{r}
## false negatives
fn <- sum(p[901:1000] > alpha)
fn
```

```{r}
## FPR = FP over all negatives
fpr <- fp / (fp+tn)

## FNR = false negative rate
fnr <- fn / (tp+fn)

## False-discovery rate = FP over all test-positives
fdr <- fp / (fp+tp)


fpr;fnr;fdr
```

The false-discovery rate (FDR) indicates the proportion of false-positives among the test positives: hence, 43% of the test-positive results are false-positives, allmost half!!

Let's write a function to compute this:

```{r}
evaluation = function(p,alpha) {
  fp <- sum(p[1:900] <= alpha)
  tn <- sum(p[1:900] > alpha)
  tp <- sum(p[901:1000] <= alpha)
  fn <- sum(p[901:1000] > alpha)
  M = matrix(c(fp,tn,tp,fn),nrow=2)
  rownames(M) = c('test pos','test neg')
  colnames(M) = c('negative','positive')
  return(M)
}

fdr = function(p,alpha) {
  fp <- sum(p[1:900] <= alpha)
  tn <- sum(p[1:900] > alpha)
  tp <- sum(p[901:1000] <= alpha)
  fn <- sum(p[901:1000] > alpha)
  fp / (fp+tp)
}
```

```{r}
evaluation(p,alpha=0.05)
```

```{r}
evaluation(p,alpha=0.01)
```

Here the FDR:

```{r}
fdr(p,alpha=0.05)
```

```{r}
fdr(p,alpha=0.01)
```

```{r}
fdr(p,alpha=0.05/1000)
```