---
title: "Plenum 13.11.25"
author: "Carl Herrmann, Maïwen Caudron-Herger"
date: "11/14/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Distributions

Find all distributions

```{r}
help(distributions)
```

### density

```{r}
x = seq(-4,4,by=0.1)
dnorm(x)
```

```{r}
plot(x,dnorm(x),type='l',lwd=3,col='red');lines(x,dnorm(x,3,2),type='l',lwd=3,col='blue')
```

### Quantiles

```{r}
#p = seq(0,1,by=0.05)
p = c(0,0.25,0.5,0.75,1)
qnorm(p,3)
```

### Cumulative distribution

```{r}
x = seq(-4,4,by=0.1)
P0 = pnorm(x)
P0
```

```{r}
P1 = pnorm(x,lower.tail = FALSE)
P1
```

```{r}
plot(x,P0,type='l',col='red',lwd=3,main='Cumulative distributions',xlab='x value',ylab='Probability');lines(x,P1,col='blue',lwd=3)
```

## Relationship between distributions: Poisson vs. Normal distribution

We have seen that in certain cases, distributions become equivalent to each other. For example the Poisson distribution becomes equivalent to the normal distribution when $\lambda\to\infty$. Let's check that!

First, we visualize the distributions

```{r}
l = 3; mu= l ; s = sqrt(l)
x0 = seq(mu-3*s,mu+3*s,by=0.01);y0 = dnorm(x0,mean=mu,sd=s)
x1 = 0:(10*l) ;y1 = dpois(x1,lambda=l);names(y1) = x1
plot(x0,y0,col='red',lwd=3,type='l',ylim=c(0,max(c(y1,y0))));points(x1,y1,pch=19,col='blue')
```

Let's make a QQplot

```{r}
p = seq(0.005,0.995,by=0.005)
x = qpois(p,lambda=l) # quantiles of the Poisson distribution
y = qnorm(p,mean=mu,sd=s) # quantiles of the normal distribution
plot(x,y,pch=20);abline(0,1)
# Alternative
qqnorm(qpois(p,lambda=l));qqline(qpois(p,lambda=l)) # Why does it look slightly different from the previous plot?
```

## Poisson vs Binomial

```{r}
lambda = 3
n = 5; p = lambda/n
x= 0:(5*lambda)
y1 = dpois(x,lambda=lambda)
y2 = dbinom(x,size=n,prob=p)
#plot(x,y1,pch=19,col='blue',ylim=c(0,max(c(y1,y2))));points(x,y2,pch=19,col='red')
plot(x,y2,pch=19,col='red');points(x,y1,pch=19,col='blue')
```

