---
title: "T-tests"
author: "Carl Herrmann"
date: "12/9/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Comparing cholesterol


```{r}
dat = read.delim('https://tinyurl.com/y4fark9g', stringsAsFactors = FALSE)
```

```{r}
chol.m = dat$chol[dat$gender=='male']
chol.f = dat$chol[dat$gender=='female']
```

```{r}
boxplot(chol.m,chol.f)
```

### Question: Do men have a higher cholesterol level compared to women?

* H0: men have a cholesterol level **lower or equal** than women
* H1: men have a cholesterol level **higher** than women

```{r}
tt = t.test(chol.m,chol.f,alternative = 'greater',var.equal = TRUE)
tt
```

Let us look at the null distribution

```{r}
df = 400
x = seq(-5,5,by=0.01)
y = dt(x,df=df)
plot(x,y,type='l',lwd=3);abline(v=tt$statistic,lwd=2,col='blue',lty=3)
```


### Question: Do men have a lower cholesterol level compared to women?

* H0: men have a cholesterol level **higher or equal** than women
* H1: men have a cholesterol level **lower** than women

```{r}
tt = t.test(chol.m,chol.f,alternative = 'less',var.equal = TRUE)
tt
```

Let us look at the null distribution

```{r}
df = 400
x = seq(-5,5,by=0.01)
y = dt(x,df=df)
plot(x,y,type='l',lwd=3);abline(v=tt$statistic,lwd=2,col='blue',lty=3)
```


### Question: Do men have a different cholesterol level compared to women?

* H0: men have a cholesterol level **equal** to that of women
* H1: men have a cholesterol level **different** than that women

```{r}
tt = t.test(chol.m,chol.f,alternative = 'two.sided',var.equal = TRUE)
tt
```

Let us look at the null distribution

```{r}
df = 400
x = seq(-5,5,by=0.01)
y = dt(x,df=df)
plot(x,y,type='l',lwd=3);abline(v=c(-abs(tt$statistic),abs(tt$statistic)),lwd=2,col='blue',lty=3)
```


## Comparing height

```{r}
height.m = dat$height[dat$gender=='male']
height.f = dat$height[dat$gender=='female']
```


### Question: Do men have a higher height compared to women?

* H0: men have a height  **lower or equal** than women
* H1: men have a height  **higher** than women

```{r}
tt = t.test(height.m,height.f,alternative = 'greater',var.equal = TRUE)
tt
```

Let us look at the null distribution

```{r}
df = 396
x = seq(-20,20,by=0.01)
y = dt(x,df=df)
plot(x,y,type='l',lwd=3);abline(v=tt$statistic,lwd=2,col='blue',lty=3)
```


### Question: Do men have a lower height  compared to women?

* H0: men have a height  **higher or equal** than women
* H1: men have a height  **lower** than women

```{r}
tt = t.test(height.m,height.f,alternative = 'less',var.equal = TRUE)
tt
```

Let us look at the null distribution

```{r}
df = 396
x = seq(-20,20,by=0.01)
y = dt(x,df=df)
plot(x,y,type='l',lwd=3);abline(v=tt$statistic,lwd=2,col='blue',lty=3)
```


### Question: Do men have a different height  compared to women?

* H0: men have a height  **equal** to that of women
* H1: men have a height  **different** than that women

```{r}
tt = t.test(height.m,height.f,alternative = 'two.sided',var.equal = TRUE)
tt
```

Let us look at the null distribution

```{r}
df = 400
x = seq(-20,20,by=0.01)
y = dt(x,df=df)
plot(x,y,type='l',lwd=3);abline(v=c(-abs(tt$statistic),abs(tt$statistic)),lwd=2,col='blue',lty=3)
```
