---
title: "Distributions"
author: "Carl Herrmann"
date: "11/24/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Binomial distribution

The binomial distribution requires **2 parameters**: the probability $p$ of a success and the number of trials $L$:

### Theoretical probability distribution

```{r}
L = 10
p = 0.2
P = dbinom(0:10,L,p)
names(P) = 0:10
barplot(P,main='Binomial distribution')
```

### Random realizations from binomial distribution

```{r}
n = 10
x = rbinom(n,L,p)
x
```

Histogram:
```{r}
barplot(table(x))
```


## Poisson distribution

The Poisson distribution depends on only one parameter, the rate $\lambda$. This rate is a real value!

### Theoretical probability distribution

```{r}
lambda = 3.1
P = dpois(0:10,lambda)
names(P) = 0:10
barplot(P,main='Poisson distribution')
```

### Random realizations from binomial distribution

```{r}
n = 100000
x = rpois(n,lambda)
x
```

Histogram:
```{r}
barplot(table(x))
```

## Negative binomial

The negative binomial distribution depends on 2 paramters: the probability of a success $p$ and the number of success to reach $r$. The negative binomial describes the waiting time until $r$ successes reached.

### Theoretical probability distribution

```{r}
r = 3
p = 1/3
P = dnbinom(0:20,size = r,prob = p)
names(P) = 0:20
barplot(P)
```

### Random realizations from negative binomial distribution

```{r}
n = 50
x = rnbinom(n,prob=p,size=r)
x
```

Histogram:
```{r}
barplot(table(x))
```

### Example of negative binomial

```{r}
## read a count table
counts = read.table('https://www.dropbox.com/s/frxd16f7xn52y7d/GTEx_top1000Genes.csv?dl=1',header=TRUE)
rownames(counts) = counts[,1]
## we remove the first 2 columns
counts = counts[,-c(1:2)]
counts[1:5,1:5]
```

We now compute the mean and the variance of each gene (row) over all samples (columns)

```{r}
m = apply(counts,1,mean)
v = apply(counts,1,var)
```

We now plot the variance vs. mean; given that the distributions are very skewed, we apply a log transformation:

```{r}
plot(log(m),log(v),pch=20,xlab='log(mean)',ylab='log(variance)');abline(0,1)
```

We observe that for lowly expressed genes, the relation between mean and variance is approximately given by $\text{variance} \sim \text{mean}$. However, for more highly expressed genes, the variance is much higher than the mean, which cannot be modelled by a Poisson or binomial distribution!


## Normal distribution


The normal or Gauss distribution is characterized by two parameters, the mean $\mu$ and standard deviation $\sigma$.

### Theoretical density distribution

```{r}
mu = 2.5
sigma = 3.1
x = seq(-6,15,by=0.1)
y = dnorm(x,mean=mu,sd=sigma)
plot(x,y,type='l',lwd=3,col='red')
```

### Random sampling from normal distribution

```{r}
n = 10000
x = rnorm(n,mean=mu,sd=sigma)
x
```

```{r}
hist(x,col='grey')
```

## t-distribution

The t-distribution describes the distribution of mean values for small samples. Because the effect of outliers in small samples is more important than in large samples, the distribution has wider tails. The t-distribution is characterized by the number of degrees of freedom, which is the sample size minus 1:

```{r}
df = 3
x = seq(-6,6,by=0.1)
y = dt(x,df=df)
plot(x,y,type='l',lwd=3,col='red')
```

Let us compare this t-distribution with a standard normal distribution!

```{r}
df = 3
x = seq(-6,6,by=0.1)
y1 = dt(x,df=df)
y2 = dnorm(x)
plot(x,y2,type='l',lwd=3,col='blue');lines(x,y1,lwd=3,col='red')
```

We see that the t-distribution (red) has larger tails compared to the standard normal distribution (blue)!